Is structure based drug design ready for selectivity optimization?

### 2018-12-19 manuscript.tex
updated methods section and early results

\paragraph{Numerical model of selectivity}
To model the impact correlation would have on the expected uncertainty for selectivity predictions, $\sigma_{selectivity}$ was calculated using Equation~\ref{eq11} for 1000 evenly spaced values of the correlation coefficient ($\rho$) from 0 to 1, for a number of combinations of per target errors ($\sigma_{target1}$ and $\sigma_{target2}$)

\begin{equation}\label{eq11}
\sigma_{selectivity} = \sqrt{\sigma_{target1}^2 + \sigma_{target2}^2 - 2\rho\sigma_{target1}\sigma_{target2}}
\end{equation}

The speed up in selectivity optimization that could be expected from using free energy calculations of a particular per target error ($\sigma_{selectivity}$) was quantify as follows using NumPy (v 1.14.2). An original, true distribution for the changed in selectivity of 200000000 new compounds proposed with respected to a reference compound was modeled as a normal distribution centered around 0 with a standard deviation of 1 kcal/mol. This assumption was made on the basis that the majority of selectivity is driven by the scaffold, and R group modifications will do little to drive changes in selectivity. The 1 kcal/mol distribution is supported by the standard deviations of the selectivity in the experimental datasets referenced in this work, which are all less than, but close, to 1 kcal/mol.

Each of these proposed compounds were "screened" by a free energy calculation technique with a per target error ($\sigma_{target}$) of 1 kcal/mol~\citep{Wang:J.Am.Chem.Soc.:2015} and a specified correlation coefficient $\rho$. A $\sigma_{selectivity}$ was calculated according to Equation~\ref{eq11}. The noise of the computational method was modeled as a normal distribution centered around 0 with a standard deviation of $\sigma_{selectivity}$ and added to the "true" change in selectivity. Any compound predicted to have an improvement in selectivity of 1.4 kcal/mol (1 log unit) would then be made and have its selectivity experimental measured. The speedup value for each value of $\rho$ is calculated as the proportion of compounds made with a true selectivity gain of 1.4 kcal/mol divided by the proportion of compounds with a 1.4 kcal/mol improvement in the original distribution, where all of the compounds were made.

Finally, this process was repeated for a 100x (2.8 kcal/mol, 2 log unit) selectivity optimization and 50 linearly spaced values of the correlation coefficient ($\rho$) between 0 and 1, for four values of $\sigma_{selectivity}$ and 40000000 compounds in the original distribution.

@@

\paragraph{Ligand Pose Generation}
Ligands were extracted from the publication entries in the BindingDB as  2D SMILES strings. 3D conformations were generated using LigPrep with OPLS3~\citep{Harder2016-zn}. Ionization state was assigned using Epik at experimental pH$\pm2$. Stereoisomers were computed by retaining any specified chiralities and varying the rest. The tautomer and ionization state with the lowest epik state penalty was selected for use in the calculation. Ligand poses were generated by first aligning to the co-crystal ligand using the Largest Common Bemis-Murcko scaffold with fuzzy matching (Schrodinger 2017-4). Ligands that were poorly aligned or failed to align were then aligned using Maximum Common Substructure (MCSS). Finally, large R-groups were allowed to sample different conformations using MM-GBSA with a common core restrained. VSGB solvation model was used with the OPLS3 forcefield. No flexible residues were defined for the ligand.

@@

\paragraph{Charge Change Free Energy Calculations} 
For ligands where a protonation state change was expected to be relevant to binding based on a small state penalty, Jaguar pKa prediction calculations~\citep{Bochevarov2013-bn} were run to identify protonation state changes with pKas within 1 log unit of the experimental pH. The predicted pKas for one ligand (Shao 12b, 7.84) was within this range. To account for this, a pKa correction was performed. For this ligand, a separate perturbation map containing ligands 12a, 12c, 12b (neutral) and 12b (charged) was run for 30ns per replica using a post-calculation Coulombic charge correction. Each charge change pertubation edge was run with a 150mM NaCl concentration. The pKa correction was performed using Equation~\ref{eq1}:

\begin{equation}\label{eq1}
\Delta\Delta G_{corrected} = \Delta\Delta G_{uncorrected} - RT\log\Bigg(\frac{10^{pK_a -pH}+1}{e^{\frac{\Delta G_{neutral} - \Delta G_{charged}}{RT}} * \big(10^{pK_a - pH}+1\big)}\Bigg)
\end{equation}

$\Delta\Delta$G for each edge in perturbation map with 12a, 12c and 12b (neutral) was updated using the correction above and merged into the final map.

@@

@@

To predict the $\Delta \Delta G_{selectivity}$ of a compound, we developed a protocol that uses a relative free energy calculation (FEP+)~\citep{Wang:J.Am.Chem.Soc.:2015} to run a map of peturbations between ligands in a congeneric series, as described in depth in the methods section. The calculation is repeated for each target of interest, with identical perturbations (edges) between each ligand (nodes). Each edge represents a relative free energy calculation that quantifies the $\Delta \Delta G$ between the ligands, or nodes. By using provided experimental data, we can convert the $\Delta \Delta G$ from each edge to a single potency value for each value against that target ($\Delta G_{target}$). From this sets of calculations, we can calculate a $\Delta \Delta G_{selectivity}$ for each ligand given two targets of interest. Previous work shows that FEP+ can achieve an accuracy ($\sigma_{target}$) of roughly 1 kcal/mol when predicting potency, which is a combination of systematic forcefield and random stastistical error~\citep{Wang:J.Am.Chem.Soc.:2015}. However, it is possible that the forcefield component of that error may fortuitously cancel when computing $\Delta \Delta G_{selectivity}$, leading to a selectivity uncertainty ($\sigma_{selectivity}$) that is lower than would be expected.

@@

To demonstrate the potential impact correlation has on the uncertainty of selectivity predictions ($\sigma_{selectivity}$) using alchemical free energy techniques, we created a simple numerical model following equation~\ref{eq11}, which takes into account each of the per target errors expected from the methodology as well as the correlation in those errors. As seen in Figure ~\ref{fig:figure-1}A, if the per target errors ($\sigma_1$ and $\sigma_2$) are the same, $\sigma_{selectivity}$ approaches 0 as the correlation coefficient ($\rho$) approaches 1. If the error for the free energy method is not the same, $\sigma_{selectivity}$ gets smaller but approaches a non-zero value as $\rho$ approaches 1. 
To quantify the expected speedup in selectivity optimization, we modeled the change in selectivity with respect to a reference compound for a number of compounds a medicinal chemist might suggest as a normal distribution centered around 0 with a standard deviation of 1 kcal/mol (Figure~\ref{fig:figure-1}B, black curve), reflecting that most proposed changes would not drive large changes in selectivity. Then, suppose that each compound is screened computationally with a method free energy methodology with a per target ($\sigma_{target}$) error of 1 kcal/mol, and all compounds predicted to have a 1.4 kcal/mol improvement in selectivity are synthesized and experimentally tested (Figure~\ref{fig:figure-1}B, colored curves). The fold-change in the proportion of compounds that are made that have a true 1.4 kcal/mol improvement in selectivty compared to the original distribution can be calculated as a surrogate for the expected speedup. For a 1.4 kcal/mol selectivity improvement threshold (1 log unit), a correlation of 0.5 gives an expected speed up of 4.1x, which can be interpreted as 4.1x fewer compounds needing to be made before achieving a 1 log unit improvement in selectivity. This process can be extended for the even more difficult proposition of achieving a 2 log unit improvement in selectivity (Figure~\ref{fig:figure-1}C), where 200-300x speedups can be expected, depending on $\sigma_{target}$ for the free energy methodology.

\begin{figure}
\begin{fullwidth}
\begin{centering}
\includegraphics[width=1.0\linewidth]{figures/figure1.png}
% Need at least one blank line for centering to work.
\end{centering}
\caption{
\label{fig:figure-1}
{\bf Free energy calculations speed up selectivity optimization} \\
({\bf A})  The effect of correlation on expected errors for predicting selectivity ($\sigma_{selectivity}$) in kcal/mol. Each curve represents a different combination of target errors ($\sigma_1$ and $\sigma_2$). 
({\bf B}) The change in selectivity for molecules proposed by medicinal chemists optimizing a lead candidate can be modeled by a normal distribution centered on 0 with a standard deviation of 1 kcal/mol (black curve). Each green curve corresponds to the distribution of compounds made after screening for a 1 log unit (1.4 kcal/mol) improvement in selectivity with a free energy methodology with a 1 kcal/mol per target error and a particular correlation. The shade region of each curve corresponds to the compounds with a real 1 log unit  improvement in selectivity. The speed up is calculated as the ratio of the percentage of compounds made with a real 1 log unit improvement to the percentage of compounds that would be expected in the original distribution.  
({\bf C}) The speedup (y-axis, log scale) expected for 100x (2 log units, 2.8 kcal/mol) selectivity optimization as a function of correlation coefficient $\rho$. Each curve corresponds to a different $\sigma_{target}$ value.  
}
\end{fullwidth}
\end{figure}

@@

\begin{figure}
\begin{fullwidth}
\begin{centering}
\includegraphics[width=1.0\linewidth]{figures/figure5.pdf}
% Need at least one blank line for centering to work.
\end{centering}
\caption{
\label{fig:figure-5}
{\bf Correlation in selectivity prediction errors can be used to accelerate selectivity optimization} \\
({\bf A}) (\emph{left}) The joint posterior distribution of the prediction errors for CDK2 (X-axis) and ERK2 (Y-axis) from the Bayesian graphical model. (\emph{right}) Speedup in selectivity optimization (Y-axis) as a function of correlation coefficient (X-axis). The posterior marginal distribution of the correlation coefficient ($\rho$) is shown in gray, while the expected speed up is shown for 100x (green curve) and 10x (yellow curve) selectivity optimization. The inserted box shows the mean and 95\% confidence interval for the correlation coefficient. 
({\bf B}) (\emph{left}) The same as above, with CDK2 (X-axis) and CDK9 (Y-axis). (\emph{right}) As above, for the CDK2/CDK9 calculations.}
\end{fullwidth}
\end{figure}

>>>>

\paragraph{Numerical model of selectivity optimization speedup}
To model the impact correlation of forcefield error would have on the expected uncertainty for selectivity predictions in the regime of infinite sampling and zero statistical error, $\sigma_{selectivity}$ was calculated using Equation~\ref{eq11} for 1000 evenly spaced values of the correlation coefficient ($\rho$) from 0 to 1, for a number of combinations of per target forcefield errors ($\sigma_{ff,1}$ and $\sigma_{ff,2}$)

\begin{equation}\label{eq11}
\sigma_{selectivity} = \sqrt{\sigma_{ff,1}^2 + \sigma_{ff,2}^2 - 2\rho\sigma_{ff,1}\sigma_{ff,2}}
\end{equation}

The speed up in selectivity optimization that could be expected from using free energy calculations of a particular per target error ($\sigma_{selectivity}$) was quantified as follows using NumPy (v 1.14.2). An original, true distribution for the changed in selectivity of 200000000 new compounds proposed with respect to a reference compound was modeled as a normal distribution centered around 0 with a standard deviation of 1 kcal/mol. This assumption was made on the basis that the majority of selectivity is driven by the scaffold, and R group modifications will do little to drive changes in selectivity. The 1 kcal/mol distribution is supported by the standard deviations of the selectivity in the experimental datasets referenced in this work, which are all less than, but close, to 1 kcal/mol.

Each of these proposed compounds were "screened" by a free energy calculation technique with a per target forcefield error ($\sigma_{ff}$) of 1 kcal/mol and a specified correlation coefficient $\rho$. A $\sigma_{selectivity}$ was calculated according to Equation~\ref{eq11}. The noise of the computational method was modeled as a normal distribution centered around 0 with a standard deviation of $\sigma_{selectivity}$ and added to the "true" change in selectivity, giving us the predicted change in selectivity ($\Delta S_{compound}$). This process can be described by Equation~\ref{eq14}:

\begin{equation}\label{eq14}
\Delta S_{compound} = \mathcal{N}_{true}(\mu =0, \sigma^2 = 1) + \mathcal{N}_{forcefield}(\mu =0, \sigma_{selectivity}^2(\rho))
\end{equation}

Any compound predicted to have an improvement in selectivity of 1.4 kcal/mol (1 log unit) would then be made and have its selectivity experimental measured, using an experimental method with perfect accuracy. The speedup value for each value of $\rho$ is calculated as the proportion of compounds made with a true selectivity gain of 1.4 kcal/mol divided by the proportion of compounds with a 1.4 kcal/mol improvement in the original distribution, where all of the compounds were made.

This process was repeated for a 100x (2.8 kcal/mol, 2 log unit) selectivity optimization and 50 linearly spaced values of the correlation coefficient ($\rho$) between 0 and 1, for four values of $\sigma_{selectivity}$ and 40000000 compounds in the original distribution.

\paragraph{Numerical model of impact of statistical error on selectivity optimization}
To model the impact of statistical error on selectivity optimization at different levels of correlation in the forcefield error, a similar scheme as above was used. An original, true distribution of 40000000 compounds was proposed with respect to a reference compound, drawing from a normal distribution centered on 0 with a standard deviation of 1 (Numpy v 1.14.2). Each of these proposed compounds were "screened" by a free energy calculation technique with a per target forcefield error ($\sigma_{ff}$) of 0.9 kcal/mol~\citep{Harder2016-zn} and a specified correlation coefficient $\rho$, which was evenly spaced between 0 and 1 in 50 steps.  A $\sigma_{selectivity}$ was calculated according to Equation~\ref{eq11}. Additionally, a per target statistical error ($\sigma_{stat}$) was as in Equation~\ref{eq15}

\begin{equation}\label{eq15}
\sigma_{stat} = \sqrt{\frac{2\sigma^2}{N}}
\end{equation}

Where N is the effort put into running sampling the calculation and $\sigma$ is such that when N is 1, $\sigma_{stat}$ = 0.2 kcal/mol. The statistical error was propagated assuming it was uncorrelated, such as in Equation~\ref{eq11} where $\rho$ = 0, giving us $\sigma_{statistics}$ The forcefield and statistical errors where modeled as Gaussian noise added to the true distribution, as in Equation~\ref{eq16}.

 \begin{equation}\label{eq16}
 \Delta S_{compound} = \mathcal{N}_{true}(\mu =0, \sigma^2 = 1) + \mathcal{N}_{forcefield}(\mu =0, \sigma_{selectivity}^2(\rho)) + \mathcal{N}_{stat}(\mu =0, \sigma_{statistics}^2(\rho))
 \end{equation}

 Any compound predicted to have an improvement in selectivity of 2.8 kcal/mol (2 log units) would then be made and have its selectivity experimental measured, using an experimental method with perfect accuracy. The speedup value for each value of $\rho$ is calculated as the proportion of compounds made with a true selectivity gain of 2.8 kcal/mol divided by the proportion of compounds with a 2.8 kcal/mol improvement in the original distribution, where all of the compounds were made.

@@

\paragraph{Ligand Pose Generation}
Ligands were extracted from the publication entries in the BindingDB as 2D SMILES strings. 3D conformations were generated using LigPrep with OPLS3~\citep{Harder2016-zn}. Ionization state was assigned using Epik at experimental pH$\pm2$. Stereoisomers were computed by retaining any specified chiralities and varying the rest. The tautomer and ionization state with the lowest epik state penalty was selected for use in the calculation. Any ligands predicted to have a positive or negative charge in its lowest Epik state penalty was excluded, with the exception of Compound 9 from the Blake dataset. This ligand was predicted to have a +1 charge for its lowest state penalty state. The neutral form the ligand was include for the sake of cycle closure in the FEP+ map, but was ignored for the sake any analysis afterwards. Ligand poses were generated by first aligning to the co-crystal ligand using the Largest Common Bemis-Murcko scaffold with fuzzy matching (Schrodinger 2017-4). Ligands that were poorly aligned or failed to align were then aligned using Maximum Common Substructure (MCSS). Finally, large R-groups were allowed to sample different conformations using MM-GBSA with a common core restrained. VSGB solvation model was used with the OPLS3 forcefield. No flexible residues were defined for the ligand.

@@

@@

 \paragraph{Calculating the marginal distribution of speedup}

 To quantify the expected speedup from the calculations we ran, we utilized 10000 replicates of the scheme detailed above to calculate speedup given parameters $\rho$, $\sigma_{ff,1}$ and $\sigma_{ff,2}$, in the regime of infinite effort and zero statistical error. Using Numpy (v 1.14.2), $\rho$ was drawn from a normal distribution with the mean and standard deviation from the posterior distribution of $\rho$ from the Bayesian Graphical model. The per target forcefield errors, $\sigma_{ff,1}$ and $\sigma_{ff,2}$, were estimated from the mean of the absolute value of  $\epsilon_{target1}$ and $\epsilon_{target2}$, which are the magnitude of errors from the Bayesian graphical model. $\sigma_{selectivity}$ was calculated using Equation~\ref{eq11}. 100000 molecules were proposed from true normal distribution, as above. The error of the computational method was modeled as in Equation~\ref{eq14}.

@@

To predict the $\Delta \Delta G_{selectivity}$ of a compound, we developed a protocol that uses a relative free energy calculation (FEP+)~\citep{Wang:J.Am.Chem.Soc.:2015} to run a map of perturbations between ligands in a congeneric series, as described in depth in the methods section. The calculation is repeated for each target of interest, with identical perturbations (edges) between each ligand (nodes). Each edge represents a relative free energy calculation that quantifies the $\Delta \Delta G$ between the ligands, or nodes. By using provided experimental data, we can convert the $\Delta \Delta G$ from each edge to a single potency value for each value against that target ($\Delta G_{target}$). From this sets of calculations, we can calculate a $\Delta \Delta G_{selectivity}$ for each ligand given two targets of interest. Previous work shows that FEP+ can achieve an accuracy ($\sigma_{target}$) of roughly 1 kcal/mol when predicting potency, which is a combination of systematic forcefield and random statistical error~\citep{Wang:J.Am.Chem.Soc.:2015}. However, it is possible that the forcefield component of that error ($\sigma_{ff}$) may fortuitously cancel when computing $\Delta \Delta G_{selectivity}$, leading to a forcefield component of the selectivity uncertainty that is lower than would be expected.

@@

To demonstrate the potential impact correlation has on the forcefield error of selectivity predictions ($\sigma_{selectivity}$) using alchemical free energy techniques, we created a simple numerical model following equation~\ref{eq11}, which takes into account each of the per target forcefield errors expected from the methodology as well as the correlation in those errors. As seen in Figure ~\ref{fig:figure-1}A, if the per target forcefield errors ($\sigma_{ff,1}$ and $\sigma_{ff,2}$) are the same, $\sigma_{selectivity}$ approaches 0 as the correlation coefficient ($\rho$) approaches 1. If the error for the free energy method is not the same, $\sigma_{selectivity}$ gets smaller but approaches a non-zero value as $\rho$ approaches 1. 
To quantify the expected speedup in selectivity optimization, we modeled the change in selectivity with respect to a reference compound for a number of compounds a medicinal chemist might suggest as a normal distribution centered around 0 with a standard deviation of 1 kcal/mol (Figure~\ref{fig:figure-1}B, black curve), reflecting that most proposed modifications would not drive large changes in selectivity. Then, suppose that each compound is screened computationally with a method free energy methodology with a per target forcefield error ($\sigma_{ff}$) of 1 kcal/mol in the regime of infinite computation effort where statistical error is 0 kcal/mol. All compounds predicted to have a 1.4 kcal/mol improvement in selectivity are synthesized and experimentally tested (Figure~\ref{fig:figure-1}B, colored curves), using an experimental technique with perfect accuracy. The fold-change in the proportion of compounds that are made that have a true 1.4 kcal/mol improvement in selectivity compared to the original distribution can be calculated as a surrogate for the expected speedup. For a 1.4 kcal/mol selectivity improvement threshold (1 log unit), a correlation of 0.5 gives an expected speed up of 4.1x, which can be interpreted as 4.1x fewer compounds needing to be made before achieving a 1 log unit improvement in selectivity. This process can be extended for the even more difficult proposition of achieving a 2 log unit improvement in selectivity (Figure~\ref{fig:figure-1}C), where 200-300x speedups can be expected, depending on $\sigma_{ff}$ for the free energy methodology.

\begin{figure}
\begin{fullwidth}
\begin{centering}
\includegraphics[width=1.0\linewidth]{figures/figure1.png}
% Need at least one blank line for centering to work.
\end{centering}
\caption{
\label{fig:figure-1}
{\bf Free energy calculations speed up selectivity optimization} \\
({\bf A})  The effect of correlation on expected errors for predicting selectivity ($\sigma_{selectivity}$) in kcal/mol. Each curve represents a different combination of per target forcefield errors ($\sigma_{ff,1}$ and $\sigma_{ff,2}$). 
({\bf B}) The change in selectivity for molecules proposed by medicinal chemists optimizing a lead candidate can be modeled by a normal distribution centered on 0 with a standard deviation of 1 kcal/mol (black curve). Each green curve corresponds to the distribution of compounds made after screening for a 1 log unit (1.4 kcal/mol) improvement in selectivity with a free energy methodology with a 1 kcal/mol per target forcefield error and a particular correlation, in the regime of infinite error where statistical error is zero. The shaded region of each curve corresponds to the compounds with a real 1 log unit  improvement in selectivity. The speedup is calculated as the ratio of the percentage of compounds made with a real 1 log unit improvement to the percentage of compounds that would be expected in the original distribution.  
({\bf C}) The speedup (y-axis, log scale) expected for 100x (2 log units, 2.8 kcal/mol) selectivity optimization as a function of correlation coefficient $\rho$. Each curve corresponds to a different $\sigma_{ff}$ value.  
}
\end{fullwidth}
\end{figure}

@@

\begin{figure}
\begin{fullwidth}
\begin{centering}
\includegraphics[width=1.0\linewidth]{figures/figure5.pdf}
% Need at least one blank line for centering to work.
\end{centering}
\caption{
\label{fig:figure-5}
{\bf Correlation in selectivity prediction errors can be used to accelerate selectivity optimization} \\
({\bf A}) (\emph{left}) The joint posterior distribution of the prediction errors for CDK2 (X-axis) and ERK2 (Y-axis) from the Bayesian graphical model. (\emph{right}) Speedup in selectivity optimization (Y-axis) as a function of correlation coefficient (X-axis). The posterior marginal distribution of the correlation coefficient ($\rho$) is shown in gray, while the expected speed up is shown for 100x (green curve) and 10x (yellow curve) selectivity optimization. The inserted box shows the mean and 95\% confidence interval for the correlation coefficient. The marginal distribution of speedup is shown on the right side of the plot for both 100x (green) and 10x (yellow) selectivity optimization speedups. 
({\bf B}) (\emph{left}) The same as above, with CDK2 (X-axis) and CDK9 (Y-axis). (\emph{right}) As above, for the CDK2/CDK9 calculations.}
\end{fullwidth}
\end{figure}
